\section{Results}

\subsection{Segmentation Results}

The current benchmark state contains two complementary sources of evidence. First, 75 completed checkpoints from the 75-run Cohort 2 inpainting semantic matrix have been evaluated on the locked C3-clean test set. These are held-out test metrics and are therefore the primary evidence for current empirical claims. Second, 75 training histories are present, including 0 runs that are still active. Validation metrics are useful for monitoring training health, but they are not used as final evidence for cross-domain generalization.

\begin{table}[H]
  \centering
  \caption{Current execution status for the Cohort 2 inpainting semantic benchmark. The matrix contains five inpainting conditions, five semantic segmentation models, and three seeds.}
  \label{tab:current-status}
  \begin{tabular}{lr}
    \toprule
    \textbf{Status item} & \textbf{Count} \\
    \midrule
    Planned Cohort 2 inpainting runs & 75 \\
    C3-clean evaluated completed checkpoints & 75 \\
    Training histories present & 75 \\
    Active training runs & 0 \\
    Not yet started runs & 0 \\
    \bottomrule
  \end{tabular}
\end{table}

Table~\ref{tab:c2-condition-results} summarizes held-out C3-clean performance by inpainting condition. C2 SDXL inpaint has the strongest condition-level mean performance, with Dice 0.443 $\pm$ 0.126 and IoU 0.347 $\pm$ 0.103. All five inpainting conditions have 15 evaluated runs, so the condition-level comparison is balanced across the five semantic segmentation models and three random seeds.

\begin{table}[H]
  \centering
  \caption{Current C3-clean performance by Cohort 2 inpainting condition. Values are mean $\pm$ standard deviation over completed evaluated checkpoints only.}
  \label{tab:c2-condition-results}
  \begin{tabular}{lrrrr}
    \toprule
    \textbf{Condition} & \textbf{Runs} & \textbf{Dice} & \textbf{IoU} & \textbf{Boundary F1} \\
    \midrule
    C2 SDXL inpaint & 15 & 0.443 $\pm$ 0.126 & 0.347 $\pm$ 0.103 & 0.500 \\
    C2 SD2 inpaint & 15 & 0.421 $\pm$ 0.183 & 0.328 $\pm$ 0.147 & 0.479 \\
    C2 SDXL texture inpaint & 15 & 0.406 $\pm$ 0.153 & 0.315 $\pm$ 0.121 & 0.461 \\
    C2 SD2 fiber inpaint & 15 & 0.398 $\pm$ 0.169 & 0.308 $\pm$ 0.136 & 0.453 \\
    C2 FLUX inpaint & 15 & 0.387 $\pm$ 0.148 & 0.300 $\pm$ 0.118 & 0.439 \\
    \bottomrule
  \end{tabular}
\end{table}

Model choice has a larger effect than the current differences among completed inpainting conditions. Table~\ref{tab:c2-model-results} shows that U-Net++ EfficientNet-B4 is the strongest completed model family, with mean Dice 0.569 $\pm$ 0.056 and IoU 0.459 $\pm$ 0.050. This pattern suggests that decoder capacity, multiscale features, and model architecture remain central for small foreground segmentation under ecological clutter.

\begin{table}[H]
  \centering
  \caption{Current C3-clean performance by segmentation model. Values are mean $\pm$ standard deviation over completed evaluated checkpoints only.}
  \label{tab:c2-model-results}
  \begin{tabular}{lrrrr}
    \toprule
    \textbf{Model} & \textbf{Runs} & \textbf{Dice} & \textbf{IoU} & \textbf{Boundary F1} \\
    \midrule
    U-Net++ EfficientNet-B4 & 15 & 0.569 $\pm$ 0.056 & 0.459 $\pm$ 0.050 & 0.635 \\
    MONAI U-Net & 15 & 0.506 $\pm$ 0.043 & 0.380 $\pm$ 0.037 & 0.579 \\
    FPN EfficientNet-B3 & 15 & 0.488 $\pm$ 0.045 & 0.383 $\pm$ 0.036 & 0.557 \\
    DeepLabV3+ ResNet50 & 15 & 0.328 $\pm$ 0.074 & 0.249 $\pm$ 0.060 & 0.382 \\
    U-Net ResNet34 & 15 & 0.162 $\pm$ 0.063 & 0.126 $\pm$ 0.049 & 0.179 \\
    \bottomrule
  \end{tabular}
\end{table}

The best completed single run is C2 SD2 inpaint with U-Net++ EfficientNet-B4 at seed 37, which achieves C3-clean Dice 0.651 with a 95\% bootstrap confidence interval of 0.595--0.703 and IoU 0.531 with a 95\% bootstrap confidence interval of 0.480--0.579. These top-run results show that synthetic ecological inpainting can support useful C3-clean segmentation when paired with a strong segmentation architecture, while the spread across seeds and models cautions against relying on any single generator-model pairing.

\begin{table}[H]
  \centering
  \caption{Training-validation summary by condition. These values monitor training health and checkpoint selection; they are not substitutes for held-out C3-clean metrics.}
  \label{tab:validation-condition-results}
  \begin{tabular}{lrrrr}
    \toprule
    \textbf{Condition} & \textbf{Runs} & \textbf{Best val Dice} & \textbf{Best val IoU} & \textbf{Last val Dice} \\
    \midrule
    C2 FLUX inpaint & 15 & 0.963 $\pm$ 0.013 & 0.933 $\pm$ 0.023 & 0.961 \\
    C2 SD2 inpaint & 15 & 0.963 $\pm$ 0.013 & 0.932 $\pm$ 0.023 & 0.962 \\
    C2 SDXL texture inpaint & 15 & 0.962 $\pm$ 0.013 & 0.932 $\pm$ 0.024 & 0.961 \\
    C2 SDXL inpaint & 15 & 0.962 $\pm$ 0.013 & 0.932 $\pm$ 0.024 & 0.959 \\
    C2 SD2 fiber inpaint & 15 & 0.962 $\pm$ 0.013 & 0.932 $\pm$ 0.024 & 0.959 \\
    \bottomrule
  \end{tabular}
\end{table}

The earlier single-condition smoke validation remains useful as an end-to-end check of the pipeline. This run trained a U-Net with a ResNet34 encoder using one seed on an SDXL-inpainted Cohort 2 synthetic training condition, then evaluated the resulting checkpoint on the 97-image Cohort 3 test set.

\begin{table}[H]
  \centering
  \caption{Preliminary single-condition Cohort 3 validation metrics. This run validates the end-to-end evaluation workflow; the full benchmark is required for condition-level conclusions.}
  \label{tab:preliminary-validation}
  \begin{tabular}{lcc}
    \toprule
    \textbf{Metric} & \textbf{Mean} & \textbf{95\% bootstrap CI} \\
    \midrule
    Dice & 0.067 & 0.041--0.096 \\
    Mask IoU & 0.041 & 0.024--0.061 \\
    Precision & 0.189 & -- \\
    Recall & 0.051 & -- \\
    Boundary F1 & 0.085 & -- \\
    Foreground area error & 0.015 & -- \\
    \bottomrule
  \end{tabular}
\end{table}

Because this smoke run uses one synthetic condition, one architecture, and one seed, we interpret it only as pipeline validation rather than final evidence for a synthetic-data benefit. The current C2 benchmark tables provide stronger evidence because they aggregate multiple inpainting conditions, models, and seeds. The difference between high validation Dice values in Table~\ref{tab:validation-condition-results} and lower C3-clean Dice values in Tables~\ref{tab:c2-condition-results}--\ref{tab:c2-model-results} is scientifically important: it shows that within-condition validation performance can overstate ecological generalization. The benchmark therefore reports C3-clean metrics as the primary outcome.
